SDGs

### 2024-03-05 inequality-SDGs.Rmd
national analysis first steps

p <- dat |>
    filter(indicator_name %in% other_vars, country_code %in% countries_list) |> 
    select(-x35) |> 
    pivot_longer(cols = x1990:last_col(), values_to = "value", names_to = "year") |> 
    mutate(year = str_remove(year, "x"), year = as.numeric(year)) |> 
    group_by(indicator_name, year) |> 
    mutate(nas = is.na(value)) |> 
    summarise(missing = sum(nas) / n_countries) |> 
    ungroup() |> group_by(indicator_name) |> 
    summarize(mean_missing = mean(missing)) |> 
    filter(mean_missing < 0.3) |> 
    ggplot(aes(mean_missing, indicator_name)) +
    geom_point() +
    #scale_fill_viridis_c(
    #    guide = guide_colorbar(barwidth = unit(2,"mm"), barheight = unit(20, "mm"))) + 
    labs(x = "Mean proportion of missing values across countries", y = "Indicator name", tag = "A")+
    theme_light(base_size = 6)

@@

q <- dat |>
    filter(indicator_name %in%  vars) |>
    select(-x35) |>
    pivot_longer(cols = x1990:last_col(), values_to = "value", names_to = "year") |>
    mutate(year = str_remove(year, "x"), year = as.numeric(year)) |>
    mutate(nas = is.na(value)) |>
    filter(year >= 2000, year < 2019) |>  # restricting time improves
    group_by(country_code, indicator_name) |>
    summarise(missing = sum(nas) / diff(range(year))) |>
    #filter(country_code %in% complete_countries) |> # this line only if next cell already run
    # ungroup() |> group_by(country_code) |>
    # summarize(mean_missing = mean(missing)) |>
    #filter(mean_missing < 0.3) |>
    ggplot(aes(country_code, indicator_name)) +
    geom_tile(aes(fill = missing)) +
    scale_fill_viridis_c(
        "Proportion of missing values",
        guide = guide_colorbar(barwidth = unit(2,"cm"), barheight = unit(2,"mm"))) + 
    labs(x = "Country code", y = "Indicator name", tag = "B") +
    theme_light(base_size = 6) +
    theme(axis.text.x = element_text(angle = 90, size = 3), 
          legend.position = "bottom")

p/q
# ggsave(
#     plot = p/q, filename = "WB_selected_vars.png", path = "figures/", device = "png",
#     width = 7, height = 7, dpi = 400, bg = "white"
# )

@@

## correct Namibia is not a NA, but the country code is NA
ineq_dat <- ineq_dat |> 
    mutate(country = case_when(is.na(country) ~ "NA", TRUE ~ country)) #|> filter(country == "NA")
```
`m` is total, `a` is average. All other codes are documented here: [https://wid.world/codes-dictionary/#packages](https://wid.world/codes-dictionary/#packages).

@@

```{r warnings=FALSE, message=FALSE}
crg <- dat |>
    filter(indicator_name %in% vars, country_code %in% complete_countries) |> 
    select(-x35) |> 
    pivot_longer(cols = x1990:last_col(), values_to = "value", names_to = "year") |> 
    mutate(year = str_remove(year, "x"), year = as.numeric(year)) |> 
    filter(year >= 2000, year < 2019) |>  # restricting time improves
    select(-indicator_code) |> 
    pivot_wider(values_from = value, names_from = indicator_name) |> 
    GGally::ggpairs(
        columns = 4:12 ,
    lower = list(continuous = GGally::wrap("points", alpha = 0.5, size = 0.5)),
    diag = list(continuous = GGally::wrap("densityDiag", alpha = 0.2))
    # upper = list(continuous = GGally::wrap("cor", size = 2))
    ) + theme_light(base_size = 4)

@@

# ggsave(
#     plot = crg, filename = "WB_correlogram.png", path = "figures/", device = "png",
#     width = 7, height = 7, dpi = 400, bg = "white"
# )
```
Interpolate missing values

@@

a <- sdg |> 
    rename(indicator_name = series_description, year = time_period, iso3 =`ISO-alpha3 Code` ) |> 
    filter(iso3 %in% countries_list, !is.na(value)) |> # use only countries
    filter(goal == 10) |> 
    group_by(indicator, indicator_name, year, iso3) |> 
    summarize(mean_val = mean(value, na.rm = TRUE)) |> 
    pivot_wider(names_from = year, values_from = mean_val) |> 
    pivot_longer(cols = starts_with("2"), names_to = "year", values_to = "value",
                 names_transform = list(year = as.numeric, value = as.numeric)) |> 
    mutate(nas = is.na(value)) |> 
    group_by(indicator_name, year) |> 
    summarise(missing = sum(nas) / n_countries2) |> 
    ggplot(aes(year, indicator_name)) +
    geom_tile(aes(fill = missing)) +
    scale_fill_viridis_c(
        "Missing",
        guide = guide_colorbar(barheight = unit(2,"cm"), barwidth = unit(2,"mm"))) + 
    labs(title = "SDG10: Inequality indicators", 
         x = "Year", y = "Indicator name", tag = "A") +
    theme_light(base_size = 6)
a
# sdg |> 
#     filter(series_description == "Proportion of voting rights of developing countries in international organizations, by organization (%)")

@@

```{r}
b <- sdg |> 
    rename(indicator_name = series_description, year = time_period, iso3 =`ISO-alpha3 Code` ) |> 
    filter(iso3 %in% countries_list, !is.na(value)) |> # use only countries
    filter(goal == 14) |> 
    group_by(indicator, indicator_name, year, iso3) |> 
    summarize(mean_val = mean(value, na.rm = TRUE)) |> 
    pivot_wider(names_from = year, values_from = mean_val) |> 
    pivot_longer(cols = starts_with("2"), names_to = "year", values_to = "value",
                 names_transform = list(year = as.numeric, value = as.numeric)) |> 
    mutate(nas = is.na(value)) |> 
    group_by(indicator_name, year) |> 
    summarise(missing = sum(nas) / n_countries2) |> 
    ggplot(aes(year, indicator_name)) +
    geom_tile(aes(fill = missing)) +
     scale_fill_viridis_c(
        "Missing",
        guide = guide_colorbar(barheight = unit(2,"cm"), barwidth = unit(2,"mm"))) +
    labs(title = "SDG14: Life under water", x = "Year", y = "Indicator name", tag = "B") +
    theme_light(base_size = 6)
b
```

```{r}
c <- sdg |> 
    rename(indicator_name = series_description, year = time_period, iso3 =`ISO-alpha3 Code` ) |> 
    filter(iso3 %in% countries_list, !is.na(value)) |> # use only countries
    filter(goal == 15) |> 
    group_by(indicator, indicator_name, year, iso3) |>
    summarize(mean_val = mean(value, na.rm = TRUE)) |>
    pivot_wider(names_from = year, values_from = mean_val) |> 
    pivot_longer(cols = starts_with("2"), names_to = "year", values_to = "value",
                 names_transform = list(year = as.numeric, value = as.numeric)) |> 
    mutate(nas = is.na(value)) |> 
    group_by(indicator, indicator_name, year) |> 
    summarise(missing = sum(nas) / n_countries2) |> 
    mutate(indicator_name = str_trunc(indicator_name, width = 100, side = "right")) |>
    ggplot(aes(year, indicator_name)) +
    geom_tile(aes(fill = missing)) +
    scale_fill_viridis_c("Missing",
        guide = guide_colorbar(barheight = unit(2,"cm"), barwidth = unit(2,"mm"))) +
    labs(title = "SDG15: Life on land", x = "Year", y = "Indicator name", tag = "C") +
    theme_light(base_size = 6)
c

@@

# ggsave(
#     plot = a/b/c, filename = "UN_selected_vars.png", path = "figures/", device = "png",
#     width = 7, height = 7, dpi = 400, bg = "white"
# )

@@

```{r}
p2 <- sdg |> 
    rename(series_name = series_description, year = time_period, iso3 =`ISO-alpha3 Code` ) |> 
    filter(iso3 %in% complete_countries) |> # use only countries
    filter(goal %in% c(10, 14, 15)) |> 
    group_by(series_name, year, iso3) |>
    # this gets rid of multiple values per year
    summarize(mean_val = mean(value, na.rm = TRUE)) |> 
    # add_count() |> 
    pivot_wider(names_from = year, values_from = mean_val) |> 
    pivot_longer(cols = starts_with("2"), names_to = "year", values_to = "value",
                 names_transform = list(year = as.numeric, value = as.numeric)) |> 
    mutate(nas = is.na(value)) |> 
    ungroup() |> group_by(series_name, iso3) |> 
   # group_by(indicator, indicator_name, iso3) |> 
    summarise(missing = sum(nas) / diff(range(year)) ) |>  ## dividing by number of years (per country)
    #pull(missing) |> range() # should be between 0-1
    # summarize(mean_missing = mean(missing)) |> 
    filter(missing < 0.3) |> 
    ungroup() |> 
    pivot_wider(names_from = series_name, values_from = missing) |> 
    pivot_longer(cols = `Average proportion of Freshwater Key Biodiversity Areas (KBAs) covered by protected areas (%)`:last_col(), names_to = "series_name",
                 values_to = "missing") |> 
    group_by(series_name) |> 
    #filter(!any(is.na(missing))) |> 
    # ggplot(aes(iso3, series_name)) +
    # geom_tile(aes(fill = missing)) +
    # theme_light()
    summarize(missing_countries = sum(is.na(missing))) |> 
    filter(missing_countries < 45) |> 
    ggplot(aes(missing_countries, series_name))+
    geom_point() +
    labs(tag = "A", y = "Series name", x = "Proportion of missing countries") +
    theme_light(base_size = 6)
p2
vars2 <- p2$data$series_name
```

```{r}
q2 <- sdg |> 
    rename(series_name = series_description, year = time_period, iso3 =`ISO-alpha3 Code` ) |> 
    filter(iso3 %in% complete_countries) |> # use only countries
    filter(series_name %in% vars2) |> 
    ## this line is to get rid of multiple values per year
    group_by(series_name, year, iso3) |>
    summarize(mean_val = mean(value, na.rm = TRUE)) |> 
    ungroup() |> #group_by(indicator_name, year, iso3) |> 
    # add_count() |> 
    pivot_wider(names_from = year, values_from = mean_val) |> 
    pivot_longer(cols = starts_with("2"), names_to = "year", values_to = "value",
                 names_transform = list(year = as.numeric, value = as.numeric)) |> 
    mutate(nas = is.na(value)) |> 
    group_by(series_name, iso3) |> 
    summarise(missing = sum(nas) / diff(range(year))) |>  ## dividing by number of years (per country)
    # summarize(mean_missing = mean(missing)) |> 
    ungroup() |> 
    filter(missing < 0.3) |> 
    ## add missing values again for combination of indicator country:
    ungroup() |> group_by(series_name) |>
    pivot_wider(names_from = iso3, values_from = missing) |> 
    pivot_longer(cols = AFG:last_col(), names_to = "iso3", values_to = "missing") |>
    ## completely empty time series
    #summarize(absent_countries = sum(is.na(missing))) |> 
    ungroup() |> group_by(iso3 ) |> 
    #filter(sum(is.na(missing)) < 45) |> # remove absent_countries > 45
    filter(!any(is.na(missing))) |> 
    ggplot(aes(iso3, series_name)) +
    #geom_point()
    geom_tile(aes(fill = missing)) +
    scale_fill_viridis_c("Missing",
        guide = guide_colorbar(barwidth = unit(2,"mm"), barheight = unit(2,"cm"))) +
    labs(tag = "B", y = "Series name", x = "Countries") +
    theme_light(base_size = 6) +
    theme(axis.text.x = element_text(angle = 90))

@@

# ggsave(
#     plot = p2/q2, filename = "UN_reduced_vars.png", path = "figures/", device = "png",
#     width = 7, height = 7, dpi = 400, bg = "white"
# )

@@

# ggsave(
#     plot = crg, filename = "UN_correlogram.png", path = "figures/", device = "png",
#     width = 7, height = 7, dpi = 400, bg = "white"
# )
```

@@

```{r}
p <- pca1$rotation |> 
    as_tibble(rownames = "vars") |> 
    separate(vars, sep = "_", into = c("year", "variable")) |> 
    mutate(year = as.numeric(year)) |> 
    pivot_longer(cols = starts_with("PC"), names_to = "PC", values_to = "load") |> 
    mutate(PC = str_remove(PC, "PC"), PC = as.numeric(PC)) |> 
    filter(PC < 5) |> 
    ggplot(aes(year, load)) +
    geom_line(aes(color = variable)) +
    facet_wrap(~PC) +
    theme_light(base_size = 8)

# ggsave(
#     plot = p, file = "pca-wb-time.png", path = "figures/", device = "png",
#     width = 7, height = 4, bg = "white", dpi = 400
# )
p + scale_color_brewer(palette = "Set3")
```

@@

a <- pca1$x |> as_tibble() |> 
    select(PC1:PC2) |> 
    mutate(iso = df_dat |> 
               filter(!is.na(rptinc992j_p0p100)) |>
               pull(country_code) |> unique()) |> 
    ggplot(aes(PC1, PC2)) +
    geom_text(aes(label = iso), size = 2) +
    coord_equal() +
    theme_light(base_size = 6)

b <- ggplot() +
    geom_segment(
        data = pca1$rotation |>
            as_tibble(rownames = "factor") |> 
            mutate(year = str_sub(factor, start = 1L, end = 4L),
                   factor = str_remove(factor, pattern = "\\d{4}_"))  |> 
            select(year, factor, PC1:PC2),
        aes(xend = PC1, yend = PC2, color = factor, alpha = as.numeric(year)),
        x = 0, y = 0,
        arrow = arrow(length = unit(0.1,"cm")), linewidth = 0.3) +
    scale_color_brewer("Variable",palette = "Set3",
        guide = guide_legend(ncol = 2, title.position = "top")) +
    scale_alpha("Year", guide = guide_legend(ncol = 1, title.position = "top")) +
    coord_equal()+
    theme_light(base_size = 6)

@@

@@

@@

mfa1 <- df_un_reduced |> 
    select(-iso3, -country) |> 
    as.matrix(dimnames = list(df_un_reduced |> pull(iso3) |> unique())) |> 
    FactoMineR::MFA(group = rep(19, 9), type = rep("s", 9)) # 19 years times 9 variables
```

@@

```
```{r}
mfa1$group
```

@@

mfa2 <- df_dat |> 
    select(-country_name, -country, -sptinc992j_p0p100) |> 
    filter(!is.na(rptinc992j_p0p100)) |> 
    pivot_longer(cols = `Access to electricity (% of population)`:last_col(),
                 values_to = "value", names_to = "variable") |> 
    arrange(variable, year) |> 
    unite(col = "var", year, variable, sep = "_") |> 
    pivot_wider(names_from = var, values_from = value) |> 
    ungroup() |> 
    select(-country_code) |> 
    as.matrix(dimnames = list(df_dat |> pull(country_code) |> unique())) |> 
    FactoMineR::MFA(group = rep(19, 12), type = rep("s", 12)) # 19 years times 9 variables
```

@@

```{r}
mfa1 |> summary()
mfa2$summary.quanti
```

```{r}
# save(pca1, pca2, mfa1, mfa2, file = "data/ordination_results.Rda")
```

>>>>

p <- dat |>
    filter(indicator_name %in% other_vars, country_code %in% countries_list) |> 
    select(-x35) |> 
    pivot_longer(cols = x1990:last_col(), values_to = "value", names_to = "year") |> 
    mutate(year = str_remove(year, "x"), year = as.numeric(year)) |> 
    group_by(indicator_name, year) |> 
    mutate(nas = is.na(value)) |> 
    summarise(missing = sum(nas) / n_countries) |> 
    ungroup() |> group_by(indicator_name) |> 
    summarize(mean_missing = mean(missing)) |> 
    filter(mean_missing < 0.3) |> 
    ggplot(aes(mean_missing, indicator_name)) +
    geom_point() +
    #scale_fill_viridis_c(
    #    guide = guide_colorbar(barwidth = unit(2,"mm"), barheight = unit(20, "mm"))) + 
    labs(x = "Mean proportion of missing values across countries", y = "Indicator name", tag = "A")+
    theme_light(base_size = 8)

@@

q <- dat |>
    filter(indicator_name %in%  vars) |>
    select(-x35) |>
    pivot_longer(cols = x1990:last_col(), values_to = "value", names_to = "year") |>
    mutate(year = str_remove(year, "x"), year = as.numeric(year)) |>
    mutate(nas = is.na(value)) |>
    filter(year >= 2000, year < 2019) |>  # restricting time improves
    group_by(country_code, indicator_name) |>
    summarise(missing = sum(nas) / diff(range(year))) |>
    #filter(country_code %in% complete_countries) |> # this line only if next cell already run
    # ungroup() |> group_by(country_code) |>
    # summarize(mean_missing = mean(missing)) |>
    #filter(mean_missing < 0.3) |>
    ggplot(aes(country_code, indicator_name)) +
    geom_tile(aes(fill = missing)) +
    scale_fill_viridis_c(
        "Proportion of missing values",
        guide = guide_colorbar(barwidth = unit(2,"cm"), barheight = unit(5,"mm"))) + 
    labs(x = "Country code", y = "Indicator name", tag = "B") +
    theme_light(base_size = 8) +
    theme(axis.text.x = element_text(angle = 90, size = 3), 
          legend.position = "bottom")

p/q
# ggsave(
#     plot = p/q, filename = "WB_selected_vars.png", path = "paper/figures/", device = "png",
#     width = 7.5, height = 7, dpi = 400, bg = "white"
# )

@@

## correct Namibia is not a NA, but the country code is NA
ineq_dat <- ineq_dat |> 
    mutate(country = case_when(is.na(country) ~ "NA", TRUE ~ country)) #|> filter(country == "NA")
```
`m` is total, `a` is average. All other codes are documented here: [https://wid.world/codes-dictionary/#packages](https://wid.world/codes-dictionary/#packages). The selected variables are:

- `rptinc992j_p0p100` which means the ratio of the pretax national income adults (pop > 20yrs) of the top 10% / the bottom 50%

- `sptinc992j_p99p100`: share of pretax national income adults equal-split for the top 1%

- `gptinc992j_p0p100`: Gini coefficient for the pretax national income for the adult population

@@

```{r warnings=FALSE, message=FALSE}
crg <- dat |>
    filter(indicator_name %in% vars, country_code %in% complete_countries) |> 
    select(-x35) |> 
    pivot_longer(cols = x1990:last_col(), values_to = "value", names_to = "year") |> 
    mutate(year = str_remove(year, "x"), year = as.numeric(year)) |> 
    filter(year >= 2000, year < 2019) |>  # restricting time improves
    select(-indicator_code) |> 
    pivot_wider(values_from = value, names_from = indicator_name) |> 
    GGally::ggpairs(
        columns = 4:12 ,
    lower = list(continuous = GGally::wrap("points", alpha = 0.5, size = 0.5)),
    diag = list(continuous = GGally::wrap("densityDiag", alpha = 0.2))
    # upper = list(continuous = GGally::wrap("cor", size = 2))
    ) + theme_light(base_size = 8)

@@

# ggsave(
#     plot = crg, filename = "WB_correlogram.png", path = "paper/figures/", device = "png",
#     width = 7, height = 7, dpi = 400, bg = "white"
# )
```
Interpolate missing values

@@

a <- sdg |> 
    rename(indicator_name = series_description, year = time_period, iso3 =`ISO-alpha3 Code` ) |> 
    filter(iso3 %in% countries_list, !is.na(value)) |> # use only countries
    filter(goal == 10) |> 
    group_by(indicator, indicator_name, year, iso3) |> 
    summarize(mean_val = mean(value, na.rm = TRUE)) |> 
    pivot_wider(names_from = year, values_from = mean_val) |> 
    pivot_longer(cols = starts_with("2"), names_to = "year", values_to = "value",
                 names_transform = list(year = as.numeric, value = as.numeric)) |> 
    mutate(nas = is.na(value)) |> 
    group_by(indicator_name, year) |> 
    summarise(missing = sum(nas) / n_countries2) |> 
    ggplot(aes(year, indicator_name)) +
    geom_tile(aes(fill = missing)) +
    scale_fill_viridis_c(
        "Missing",
        guide = guide_colorbar(barheight = unit(2,"cm"), barwidth = unit(2,"mm"))) + 
    labs(title = "SDG10: Inequality indicators", 
         x = "Year", y = "Indicator name", tag = "A") +
    theme_light(base_size = 8)
a
# sdg |> 
#     filter(series_description == "Proportion of voting rights of developing countries in international organizations, by organization (%)")

@@

```{r}
b <- sdg |> 
    rename(indicator_name = series_description, year = time_period, iso3 =`ISO-alpha3 Code` ) |> 
    filter(iso3 %in% countries_list, !is.na(value)) |> # use only countries
    filter(goal == 14) |> 
    group_by(indicator, indicator_name, year, iso3) |> 
    summarize(mean_val = mean(value, na.rm = TRUE)) |> 
    pivot_wider(names_from = year, values_from = mean_val) |> 
    pivot_longer(cols = starts_with("2"), names_to = "year", values_to = "value",
                 names_transform = list(year = as.numeric, value = as.numeric)) |> 
    mutate(nas = is.na(value)) |> 
    group_by(indicator_name, year) |> 
    summarise(missing = sum(nas) / n_countries2) |> 
    ggplot(aes(year, indicator_name)) +
    geom_tile(aes(fill = missing)) +
     scale_fill_viridis_c(
        "Missing",
        guide = guide_colorbar(barheight = unit(2,"cm"), barwidth = unit(2,"mm"))) +
    labs(title = "SDG14: Life under water", x = "Year", y = "Indicator name", tag = "B") +
    theme_light(base_size = 8)
b
```

```{r}
c <- sdg |> 
    rename(indicator_name = series_description, year = time_period, iso3 =`ISO-alpha3 Code` ) |> 
    filter(iso3 %in% countries_list, !is.na(value)) |> # use only countries
    filter(goal == 15) |> 
    group_by(indicator, indicator_name, year, iso3) |>
    summarize(mean_val = mean(value, na.rm = TRUE)) |>
    pivot_wider(names_from = year, values_from = mean_val) |> 
    pivot_longer(cols = starts_with("2"), names_to = "year", values_to = "value",
                 names_transform = list(year = as.numeric, value = as.numeric)) |> 
    mutate(nas = is.na(value)) |> 
    group_by(indicator, indicator_name, year) |> 
    summarise(missing = sum(nas) / n_countries2) |> 
    mutate(indicator_name = str_trunc(indicator_name, width = 100, side = "right")) |>
    ggplot(aes(year, indicator_name)) +
    geom_tile(aes(fill = missing)) +
    scale_fill_viridis_c("Missing",
        guide = guide_colorbar(barheight = unit(2,"cm"), barwidth = unit(2,"mm"))) +
    labs(title = "SDG15: Life on land", x = "Year", y = "Indicator name", tag = "C") +
    theme_light(base_size = 8)
c

@@

# ggsave(
#     plot = a/b/c + plot_layout(heights = c(1,1,1.5)), 
#     filename = "UN_selected_vars.png", path = "paper/figures/", device = "png",
#     width = 7, height = 8, dpi = 400, bg = "white"
# )

@@

```{r}
p2 <- sdg |> 
    rename(series_name = series_description, year = time_period, iso3 =`ISO-alpha3 Code` ) |> 
    filter(iso3 %in% complete_countries) |> # use only countries
    filter(goal %in% c(10, 14, 15)) |> 
    group_by(series_name, year, iso3) |>
    # this gets rid of multiple values per year
    summarize(mean_val = mean(value, na.rm = TRUE)) |> 
    # add_count() |> 
    pivot_wider(names_from = year, values_from = mean_val) |> 
    pivot_longer(cols = starts_with("2"), names_to = "year", values_to = "value",
                 names_transform = list(year = as.numeric, value = as.numeric)) |> 
    mutate(nas = is.na(value)) |> 
    ungroup() |> group_by(series_name, iso3) |> 
   # group_by(indicator, indicator_name, iso3) |> 
    summarise(missing = sum(nas) / diff(range(year)) ) |>  ## dividing by number of years (per country)
    #pull(missing) |> range() # should be between 0-1
    # summarize(mean_missing = mean(missing)) |> 
    filter(missing < 0.3) |> 
    ungroup() |> 
    pivot_wider(names_from = series_name, values_from = missing) |> 
    pivot_longer(cols = `Average proportion of Freshwater Key Biodiversity Areas (KBAs) covered by protected areas (%)`:last_col(), names_to = "series_name",
                 values_to = "missing") |> 
    group_by(series_name) |> 
    #filter(!any(is.na(missing))) |> 
    # ggplot(aes(iso3, series_name)) +
    # geom_tile(aes(fill = missing)) +
    # theme_light()
    summarize(missing_countries = sum(is.na(missing))) |> 
    filter(missing_countries < 45) |> 
    mutate(series_name = str_replace(series_name, " covered", "\ncovered"),
           series_name = str_replace(series_name, " certification", "\ncertification")) |> 
    ggplot(aes(missing_countries, series_name))+
    geom_point() +
    labs(tag = "A", y = "Series name", x = "Proportion of missing countries") +
    theme_light(base_size = 8)
p2
vars2 <- p2$data$series_name
```

```{r}
q2 <- sdg |> 
    rename(series_name = series_description, year = time_period, iso3 =`ISO-alpha3 Code` ) |> 
    filter(iso3 %in% complete_countries) |> # use only countries
    filter(series_name %in% vars2) |> 
    ## this line is to get rid of multiple values per year
    group_by(series_name, year, iso3) |>
    summarize(mean_val = mean(value, na.rm = TRUE)) |> 
    ungroup() |> #group_by(indicator_name, year, iso3) |> 
    # add_count() |> 
    pivot_wider(names_from = year, values_from = mean_val) |> 
    pivot_longer(cols = starts_with("2"), names_to = "year", values_to = "value",
                 names_transform = list(year = as.numeric, value = as.numeric)) |> 
    mutate(nas = is.na(value)) |> 
    group_by(series_name, iso3) |> 
    summarise(missing = sum(nas) / diff(range(year))) |>  ## dividing by number of years (per country)
    # summarize(mean_missing = mean(missing)) |> 
    ungroup() |> 
    filter(missing < 0.3) |> 
    ## add missing values again for combination of indicator country:
    ungroup() |> group_by(series_name) |>
    pivot_wider(names_from = iso3, values_from = missing) |> 
    pivot_longer(cols = AFG:last_col(), names_to = "iso3", values_to = "missing") |>
    ## completely empty time series
    #summarize(absent_countries = sum(is.na(missing))) |> 
    ungroup() |> group_by(iso3 ) |> 
    #filter(sum(is.na(missing)) < 45) |> # remove absent_countries > 45
    filter(!any(is.na(missing))) |> 
    mutate(series_name = str_replace(series_name, " covered", "\ncovered"),
           series_name = str_replace(series_name, " certification", "\ncertification")) |> 
    ggplot(aes(iso3, series_name)) +
    #geom_point()
    geom_tile(aes(fill = missing)) +
    scale_fill_viridis_c("Missing",
        guide = guide_colorbar(barwidth = unit(2,"mm"), barheight = unit(2,"cm"))) +
    labs(tag = "B", y = "Series name", x = "Countries") +
    theme_light(base_size = 8) +
    theme(axis.text.x = element_text(angle = 90, size = 4))

@@

# ggsave(
#     plot = p2/q2, filename = "UN_reduced_vars.png", 
#     path = "paper/figures/", device = "png",
#     width = 7, height = 7.5, dpi = 400, bg = "white"
# )

@@

# ggsave(
#     plot = crg, filename = "UN_correlogram.png", path = "paper/figures/", device = "png",
#     width = 7, height = 7, dpi = 400, bg = "white"
# )
```

@@

```{r}
p <- pca1$rotation |> 
    as_tibble(rownames = "vars") |> 
    separate(vars, sep = "_", into = c("year", "variable")) |> 
    mutate(year = as.numeric(year)) |> 
    pivot_longer(cols = starts_with("PC"), names_to = "PC", values_to = "load") |> 
    mutate(PC = str_remove(PC, "PC"), PC = as.numeric(PC)) |> 
    filter(PC < 5) |> 
    ggplot(aes(year, load)) +
    geom_line(aes(color = variable)) +
    labs(y = "Component loadings", x = "Year", title = "Top 4 principal components",
         caption = "Data source: World Bank and World Inequality Database") +
    facet_wrap(~PC) + scale_color_brewer("Variables", palette = "Set3") +
    theme_light(base_size = 10)
p
# ggsave(
#     plot = p, file = "pca-wb-time.png", path = "paper/figures/", device = "png",
#     width = 7, height = 4, bg = "white", dpi = 400
# )

@@

ggplot(data = x |> filter(var == "Proportion of Variance"), aes(comp, value)) +
    geom_col(color = "grey25") +
    geom_point(data = x |> filter(var == "Cumulative Proportion"), aes(comp, value)) +
    geom_line(data = x |> filter(var == "Cumulative Proportion"), aes(comp, value)) +
    labs(x = "Principal components", y = "Explained variance") +
    scale_x_continuous(n.breaks = 10) +
    scale_y_continuous(labels = scales::label_percent()) +
    theme_light(base_size = 10)

a <- pca1$x |> as_tibble() |> 
    select(PC1:PC2) |> 
    mutate(iso = df_dat |> 
               filter(!is.na(rptinc992j_p0p100)) |>
               pull(country_code) |> unique()) |> 
    ggplot(aes(PC1, PC2)) +
    geom_text(aes(label = iso), size = 2) +
    coord_equal() +
    theme_light(base_size = 10)

b <- ggplot() +
    geom_segment(
        data = pca1$rotation |>
            as_tibble(rownames = "factor") |> 
            mutate(year = str_sub(factor, start = 1L, end = 4L),
                   factor = str_remove(factor, pattern = "\\d{4}_"))  |> 
            select(year, factor, PC1:PC2),
        aes(xend = PC1, yend = PC2, color = factor, alpha = as.numeric(year)),
        x = 0, y = 0,
        arrow = arrow(length = unit(0.1,"cm")), linewidth = 0.3) +
    scale_color_brewer("Variable",palette = "Set3",
        guide = guide_legend(ncol = 2, title.position = "top")) +
    scale_alpha("Year", guide = guide_legend(ncol = 1, title.position = "top")) +
    coord_equal()+
    theme_light(base_size = 10)

@@

# ggsave(
#     filename = "pca_wb-data.png", path = "paper/figures/", dpi = 400, bg = "white",
#     device = "png", width = 7, height = 4,
#     plot = (a + b + plot_layout(guides = "collect") &
#     theme_light(base_size = 6) &
#     theme(legend.position = "bottom") )
# )

@@

**240220** To do: code categorical variables so they appear on the order of Inequality, life on land, life on water. Do the clustering routine on the PCA coordinates, use all components, or the first 10. Show the clusters on fig A.

@@

mfa1 <- df_un_reduced |> 
    select(-iso3, -country) |> 
    as.matrix(dimnames = list(df_un_reduced |> pull(iso3) |> unique())) |> 
    FactoMineR::MFA(group = rep(19, 9), type = rep("s", 9), ncp = 20) # 19 years times 9 variables, 20 components (PCAs)
```

@@

```
```{r}
#mfa1$group
```

@@

mfa2 <- df_dat |> 
    select(-country_name, -country, -sptinc992j_p0p100) |> 
    filter(!is.na(rptinc992j_p0p100)) |> 
    pivot_longer(cols = `Access to electricity (% of population)`:last_col(),
                 values_to = "value", names_to = "variable") |> 
    arrange(variable, year) |> 
    unite(col = "var", year, variable, sep = "_") |> 
    pivot_wider(names_from = var, values_from = value) |> 
    ungroup() |> 
    select(-country_code) |> 
    as.matrix(dimnames = list(df_dat |> pull(country_code) |> unique())) |> 
    FactoMineR::MFA(group = rep(19, 12), type = rep("s", 12), ncp = 20) # 19 years times 9 variables
```

@@

```{r}
# mfa1 |> summary()
# mfa2$summary.quanti
```

```{r}
#save(pca1, pca2, mfa1, mfa2, file = "data/ordination_results.Rda")
```

```{r}
df_un_reduced |> 
    pivot_longer(3:last_col(), names_to = "variables", values_to = "values") |> 
    separate(variables, sep = "_", into = c("year", "variable")) |> 
    mutate(year= as.numeric(year)) |> 
    filter(variable == "gptinc992j") |> 
    ggplot(aes(year, values)) +
    geom_line(aes(group = iso3))
```

df_un_reduced |> 
    pivot_longer(3:last_col(), names_to = "variables", values_to = "values") |> 
    separate(variables, sep = "_", into = c("year", "variable")) |> 
    mutate(year= as.numeric(year)) |> #pull(variable) |> unique()
    mutate(variable = case_when(
        variable == "Average proportion of Freshwater Key Biodiversity Areas (KBAs) covered by protected areas (%)" ~ "freshwater",
        variable == "Average proportion of Marine Key Biodiversity Areas (KBAs) covered by protected areas (%)"~"marine",
        variable == "Average proportion of Mountain Key Biodiversity Areas (KBAs) covered by protected areas (%)" ~ "mountain",
        variable == "Average proportion of Terrestrial Key Biodiversity Areas (KBAs) covered by protected areas (%)" ~ "terrestrial", 
        variable == "Forest area certified under an independently verified certification scheme (thousands of hectares)" ~ "forest",
        variable == "gptinc992j" ~ "Gini",
        variable == "rptinc992j"~ "Ratio 10/50",
        variable == "sptinc992j" ~ "Share 1%",
        .default = variable
    )) |> 
    filter(variable %in% c("freshwater", "marine", "terrestrial", "mountain", "forest", "Gini", "Ratio 10/50", "Share 1%")) |> #pull(variable) |> unique()
    pivot_wider(names_from = variable, values_from = values) |> 
    ggplot(aes(Gini, freshwater)) + 
    geom_line(aes(group = iso3, color = year))
